---
title: "Good Coding Practices"
author: 
- "Olivia Beck" 
date: "August 24, 2022"
output: 
  slidy_presentation: default
  html_notebook: default
---
<style type="text/css">
  body{
  font-family: Helvetica;
  font-size: 12pt;
}
</style>



## Ethical Programming  


<center> 

"Ethical programming includes being honest about
the extent to which you can be confident that your program is
good, and adhering to quality control processes that ensure that
if a mistake has been made – everybody makes mistakes – it is
found and rectified before it causes harm."


-PERDITA STEVENS

</center>

## Good programs.... 

- do what they are supposed to do 

- are clearly written 

  - Your code must be readable 
  
  - Use comments! 

- do not cause harm 


## R Terminology 

- **R** is the programming language.

- **Rstudio** is a user interface to R that makes it easier to document your work, access nice features of R, and more.

  - You do not need Rstudio to use R, but it does make it much easier
    
- **CRAN** is a repository where the latest downloads of R (and legacy versions) are found in addition to source code for thousands of different user contributed R packages.

- **Markdown** is a lightweight markup language that you can use to add formatting elements to plaintext text documents

  - **LaTeX** is a document preparation system for high-quality typesetting. LaTeX and Markdown work seamlessly together

- **RMarkdown** is integrating R and Markdown into one document

  - You can render RMarkdown documents into pdf's, HTML, Word doc's, beamer's, slidy's, RNotebook's

- **Packages** are collections of R functions, data, and compiled code in a well-defined format.

  - Packages need to be installed only once: `install.packages()`
  
  - Some packages are installed with R automatically:  `stats`, `graphics`, `grDevices`, `utils`, `datasets`, `methods`, `base`
  
  - All other packages need to be installed 

- **Libraries** are the directories where the packages are stored on your computer

  - Libraries need to be loaded every time you need them at the beginning of your script with `library()`
  
  - Every time you install a package, you also need to load the library
  
  - There is a way to access a function inside a package without loading it with package_name::function_name (this is more advanced and we will discuss it later)
  
  
## File Extensions 

- **.R** is an R script

- **.Rmd** is a Rmarkdown document 

  - The raw Rmarkdown document will have extension .Rmd
  
  - The rendered document will have the appropriate extension 
  
 - | File Type | Extension | 
| --------- | ----------| 
| PDF | .pdf| 
| Beamer | .pdf| 
| HTML | .html| 
| Slidy | .html| 
| Rnotebook | .nb.html|
| Word doc | .docx |


## Good Coding Practices 

- Anything you name should be meaningful. This includes file names, function names, variable names, etc. 

Let's look at the `iris` data set:

```{r}
head(iris)

# Good 
dat <- iris
# get the 2 columns that have "petal" in their names 
dat.petal <- dat[ , c("Petal.Length", "Petal.Width") ] 

greater.than.1 <- function(x){
  return(x>1)
}

# Bad
x <- iris 
z <- dat$Petal.Width

f <- function(x){
  return(x>1)
}

```

- Use "<-" (assign) not "=" (equals) when assigning variable/function values 

  - In R, "<-" is used to assign variable/function names and "=" is used inside functions for arguments 

```{r}
# Good 
x <- 1:5
quants.good <- quantile(x, probs = c(0.1, 0.9))

# Bad 
x = 1:5
quants.bad = quantile(x, probs = c(0.1, 0.9))
```


- NEVER use spaces to name thing 

  - file paths/names should use "-". e.g. file-name.R
   
  - function names should use "_". e.g. `funciton_name()`
  
  - variable names should use "." e.g. `variable.name` 

- Use lower case letters when possible 

  - good: file-name.R
  
  - bad: File-Name.R

- Avoid using base R functions as variable/function names 

```{r, eval = F}
# BAD
T <- 14

c <- 10 

mean <- FALSE
```

- Liberally use comments 

```{r}
# this is a function to determine if a number is greater than 1
greater.than.1 <- function(x){
  return(x>1)
}
```


## Using Spaces 

- Place spaces around all infix operators (=, +, -, <-, etc.). The same rule applies when using = in function calls. Always put a space after a comma, and never before (just like in regular English).
  
```{r, eval = F}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
```
  
- There are 2 small exceptions:
  
1.  :, :: and ::: can't have spaces around them.
  
```{r, eval = F}
# Good
x <- 1:10
base::get

# Bad
x <- 1 : 10
base :: get
```
  
2. Place a space before left parentheses, except in a function call.

```{r, eval = F}
# Good
if (debug) do(x) #space before left parentheses
plot(x, y) #function call 

# Bad
if(debug)do(x) #need space before left parentheses
plot (x, y) #no space before "(" in function call 
```
  
- Extra spacing (i.e., more than one space in a row) is okay (but not required) if it improves alignment of equal signs or assignments (<-).

```{r, eval = F}
# Good 
list(
  total = a + b + c, 
  mean  = (a + b + c) / n
)

longer.variable.name <- 5
variable.name        <- 1

longer.variable.name <- 5
variable.name <- 1


```

- Do not place spaces around square brackets (unless there’s a comma, in which case see above).
   
```{r, eval = F}
# Good
diamonds[5, ]

# Bad
x[1,]   # Needs a space after the comma
x[1 ,]  # Space goes after comma not before
x [1, ] # No space before "["
```


## Lines of Code 

- Strive to limit your code to 80 characters per line. This fits comfortably on a printed page with a reasonably sized font. If you find yourself running out of room, this is a good indication that you should encapsulate some of the work in a separate function.

  - Use the light gray line on the left side of the console in RStudio

- An opening curly brace should never go on its own line and should always be followed by a new line. A closing curly brace should always go on its own line, unless it’s followed by else. Always indent the code inside curly braces. 

```{r, eval = F}
# Good
if (y < 0 && debug) {
  message("Y is negative")
}

if (y == 0) {
  log(x)
} else {
  y ^ x
}

# Bad

if (y < 0 && debug)
message("Y is negative") #no curly braces and message is on next line

if (y == 0) {
  log(x)
} 
else {  #else needs to be on the same line as "}"
  y ^ x
}

```
  
- It’s okay to leave very short statements on the same line:

```{r, eval = F}
if (y < 0 && debug) message("Y is negative")
```


## Indentation 

- When indenting your code, use two spaces. Never use tabs or mix tabs and spaces.

  - a tab is usually 12 spaces
  
  - in Rstudio you can set your tab button to be 2 spaces: Tools > Global Options > Code > Editing > General > Check "Insert spaces for tab" > Set tab width equal to 2 spaces
  
  - You can set up auto-indent by: Tools > Global Options > Code > Editing > General > Check "Vertically align arguments in auto-indent"

- The only exception is if a function definition runs over multiple lines. In that case, indent the second line to where the definition starts:


```{r , eval = F}
# Good 
long_function_name <- function(a = "a long argument", 
                               b = "another argument",
                               c = "another long argument") {
  # As usual code is indented by two spaces.
  return(a + b + c)
} 

if (y == 0) {
  log(x)
} else {
  y ^ x
}

# Bad 
long_function_name <- function(a = "a long argument", 
  b = "another argument",
  c = "another long argument")  {
  # As usual code is indented by two spaces.
  return(a + b + c)
} 
  
if (y == 0) {
log(x)
} else {
y ^ x
} 
  
```

## Reminders 

- Activity 1, Homework 1, Reading Quiz are due Wednesday, August 30th at noon

- Office hours on Tuesday 

- Use Canvas Discussion Board for questions 

- Use this time to do Activity or Homework 1 and ask questions! 

## Additional Readings 

- Advanced R by Hadley Wickham Ch. 22 <http://adv-r.had.co.nz/Style.html> 

- How to Write Good Programs by Perdita Stevens Ch. 2 & 3


## References 

- https://www.statmethods.net/interface/packages.html

- https://hbctraining.github.io/Intro-to-R-flipped/lessons/04_introR_packages.html

- https://www.markdownguide.org/getting-started/

- https://rmarkdown.rstudio.com

- http://adv-r.had.co.nz/Style.html




