---
title: "Presidential Practice"
subtitle: "Guided Example"
author: "Dr. Matthew Beckman"
output: html_notebook
---


```{r include=FALSE}
# clean up
rm(list = ls())

# packages
library(tidyverse)   # also loads `ggplot2` package, which includes our data
library(tibble)      # `tibble::tribble()` will help us add more presidents 
library(lubridate)   # nice functions for handling dates

# data intake
data("presidential", package = "ggplot2")

# search help documentation for `tribble` function if you're curious how this works
AddPres <- 
    tibble::tribble(~name, ~start, ~end, ~party,
                    "Roosevelt", ymd("1933-03-04"), ymd("1945-04-12"), "Democratic",
                     "Truman", ymd("1945-04-12"), ymd("1953-01-20"), "Democratic",
                    "Trump", ymd("2017-01-20"), ymd("2021-01-20"), "Republican", 
                    "Biden", ymd("2021-01-20"), ymd(today()), "Democratic")

# append the rows of our two "data frames"
Presidents <- bind_rows(presidential, AddPres)
```


## Guided practice (Part 1) 

1. various summary functions to meet the `Presidents`
2. calculate `duration` of presidency 
3. show all records for presidents named "Bush"
4. subset the data to include only `name`, `party`, `duration`

```{r}
#1
glimpse(Presidents)  # what's the benefit?
str(Presidents)      # ?
summary(Presidents)  # not so useful?!
View(Presidents)     # console please!

#2

Presidents<-Presidents %>% 
  mutate(duration= end-start, name = if_else(name == "Bush" & start < ymd("2000-01-01"),"Bush (Sr)", name)) %>%
  select(name,party,duration)

```


## Guided practice (Part 2)

5. which 3 presidents served shortest?
6. which 3 presidents served longest?
7. which presidents served more than 4 years, but less than 8 years?
8. investigating the power balance
   - which party has most presidents?
   - which party spent most days in power?

```{r}
#5 and 6
Presidents %>%
  arrange(duration)%>%
  head(3)
Presidents %>%
  arrange(desc(duration))%>%
  head(3)

#7
Presidents %>%
    filter(duration > 1461, duration < 2922)


#8
Presidents %>% 
  group_by(party) %>%
  summarise(count= n(), Totaldays= sum(duration))
 
```




9. add Harry Truman & Franklin D. Roosevelt, then repeat 5-8


```{r}
library(dcData)
head(Minneapolis2013)

Minneapolis2013 %>%
  group_by(First) %>%
  summarise(count=n()) %>%
  select(First) 
  

Minneapolis2013 %>%
  mutate(First= if_else(First %in% minorcand$First,"minor",First),Second= if_else(Second %in% minorcand$First,"minor",Second) )


```


